---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# chemr

Chemr provides data structures for elements, molecules, and reactions, to provide a framework for chemical modelling and analysis in R. The text-based formats for elements, reactions, and molecules broadly correspond to that used by the [PHREEQC](https://wwwbrr.cr.usgs.gov/projects/GWC_coupled/phreeqc/) tool, an interface for which is [available in R](https://cran.r-project.org/package=phreeqc). An extension to the R library, [easyphreeqc](https://github.com/paleolimbot/easyphreeqc) is under active development.

## Installation

You can install chemr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("paleolimbot/chemr")
```

If you can load the package, everything worked!

```{r}
library(chemr)
```

## The periodic tibble

The periodic tibble is [Wikipedia's version of the periodic table](https://en.wikipedia.org/wiki/List_of_chemical_elements) in data.frame (tibble) form. You can access these data in a few ways:

```{r}
is_element("Hg")
```

```{r}
elmass("Hg")
```

```{r}
elz("Hg")
```

```{r}
elsymbol(80)
```
 
 You can also access the entire periodic tibble by typing `data(pt)`.
 
```{r}
data(pt)
pt
```

## Molecules

Molecules are a collection of counted elements (or sub-molecules) with a charge. While it's possible to create a molecule by "hand", it's much easier to use the character representation of a molecule, which is usually what you get when copy/pasting from a source.

```{r}
mol("H2O")
```

And like everything else in R, `mol` objects are vectorized, so you can serialize an entire column of molecule formulas.

```{r}
as_mol(c("H2O", "H+", "Fe(OH)3", "Ca+2"))
```

You can access the mass, charge, and elemental composition of a molecule using `mass()`, `charge()`, and `as.data.frame()` or `as.matrix()`

```{r}
m <- as_mol(c("H2O", "H+", "Fe(OH)3", "Ca+2"))
mass(m)
```

```{r}
charge(m)
```

```{r}
as.data.frame(m)
```

## Reactions

Reactions are essentially a molecule vector with coefficients (positive for the left side, negative for the right side). Similar to molecules, it's easiest to use the serialized form (conveniently, what is generally copied/pasted):

```{r}
as_reaction("2H2 + O2 = 2H2O")
```

The `is_balanced()` and `balance()` functions will happily balance these for you, provided you have the correct number of species defined.

```{r}
balance("H2 + O2 = H2O")
```

You can access various components of a reaction in the same way as for molecules:

```{r}
r <- as_reaction("2H2 + O2 = 2H2O")
lhs(r)
```

```{r}
rhs(r)
```

```{r}
mass(r) # mass balance of the reaction
```

```{r}
charge(r) # charge balance of the reaction
```

```{r}
as.data.frame(r)
```

```{r}
as.matrix(r)
```

## Molecule and Reaction arithmetic

Various arithmetic operators are available for molecule and reaction objects, such as `+`, `*` and `==`.

```{r}
m <- mol(~Fe2O3, ~H2O, ~NH3, ~`H+`)
m + as_mol("OH-")
```

```{r}
m * 2
```

```{r}
m == as_mol(~H2O)
```

Reactions have similar arithmetic, with coefficients to various molecules being added together.

```{r}
r1 <- as_reaction("2H2 + O2 = 2H2O")
r1 + as_reaction("H2O = H2O")
```

By default the reaction isn't simplified, but can be using `simplify()` and `remove_zero_counts()`.

```{r}
simplify(r1 + as_reaction("H2O = H2O"))
```

```{r}
simplify(r1 - as_reaction("2H+ + 2OH- = 2H2O"))
```

```{r}
remove_zero_counts(simplify(r1 - as_reaction("2H+ + 2OH- = 2H2O")))
```

## The Wish List

There are lots of things missing from this package that should exist, including the include various parameters to molecules and equations such as $\Delta H$ or aliases (e.g., `CaSO4` as "gypsum"). Additionally, there is currently no way to indicate hydration in the same way as PHREEQC (e.g., `CaSO4:2H2O`). Currently this is possible only as `CaSO4(H2O)2`. Feel free to [contribute to development](https://github.com/paleolimbot/chemr) or [submit feature requests](https://github.com/paleolimbot/chemr/issues) on GitHub.
